---
output: github_document
bibliography: "inst/REFERENCES.bib"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R/`cvCovEst`

<!-- badges: start -->
[![CircleCI](https://dl.circleci.com/status-badge/img/gh/PhilBoileau/cvCovEst/tree/master.svg?style=svg)](https://dl.circleci.com/status-badge/redirect/gh/PhilBoileau/cvCovEst/tree/master)
[![codecov](https://codecov.io/gh/PhilBoileau/cvCovEst/branch/master/graph/badge.svg?token=miHiqpGXxJ)](https://app.codecov.io/gh/PhilBoileau/cvCovEst)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![DOI](https://joss.theoj.org/papers/10.21105/joss.03273/status.svg)](https://doi.org/10.21105/joss.03273)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

> Cross-Validated Covariance Matrix Estimation

__Authors:__ [Philippe Boileau](https://pboileau.ca),
[Brian Collica](https://www.linkedin.com/in/brian-collica-553b0b94), and
[Nima Hejazi](https://nimahejazi.org)

---

## What's `cvCovEst`?

`cvCovEst` implements an efficient cross-validated procedure for covariance
matrix estimation, particularly useful in high-dimensional settings. The
general methodology allows for cross-validation to be used to data adaptively
identify the optimal estimator of the covariance matrix from a prespecified set
of candidate estimators. An overview of the framework is provided in the
package vignette. For a more detailed description, see @boileau2021. A suite of
plotting and diagnostic tools are also included.

---

## Installation

For standard use, install `cvCovEst` from
[CRAN](https://cran.r-project.org/package=cvCovEst):

```{r CRAN-install, eval=FALSE}
install.packages("cvCovEst")
```

The _development version_ of the package may be installed from GitHub using
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval=FALSE}
remotes::install_github("PhilBoileau/cvCovEst")
```

---

## Example

To illustrate how `cvCovEst` may be used to select an optimal covariance matrix
estimator via cross-validation, consider the following toy example:

```{r example, message=FALSE, warning=FALSE}
library(MASS)
library(cvCovEst)
set.seed(1584)

# generate a 50x50 covariance matrix with unit variances and off-diagonal
# elements equal to 0.5
Sigma <- matrix(0.5, nrow = 50, ncol = 50) + diag(0.5, nrow = 50)

# sample 50 observations from multivariate normal with mean = 0, var = Sigma
dat <- mvrnorm(n = 50, mu = rep(0, 50), Sigma = Sigma)

# run CV-selector
cv_cov_est_out <- cvCovEst(
    dat = dat,
    estimators = c(linearShrinkLWEst, denseLinearShrinkEst,
                   thresholdingEst, poetEst, sampleCovEst),
    estimator_params = list(
      thresholdingEst = list(gamma = c(0.2, 2)),
      poetEst = list(lambda = c(0.1, 0.2), k = c(1L, 2L))
    ),
    cv_loss = cvMatrixFrobeniusLoss,
    cv_scheme = "v_fold",
    v_folds = 5,
  )

# print the table of risk estimates
# NOTE: the estimated covariance matrix is accessible via the `$estimate` slot
cv_cov_est_out$risk_df
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file
an issue](https://github.com/PhilBoileau/cvCovEst/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/PhilBoileau/cvCovEst/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

Please cite the following paper when using the `cvCovEst` R software package.

```
@article{cvCovEst2021,
  doi = {10.21105/joss.03273},
  url = {https://doi.org/10.21105/joss.03273},
  year = {2021},
  publisher = {The Open Journal},
  volume = {6},
  number = {63},
  pages = {3273},
  author = {Philippe Boileau and Nima S. Hejazi and Brian Collica and Mark J. van der Laan and Sandrine Dudoit},
  title = {cvCovEst: Cross-validated covariance matrix estimator selection and evaluation in `R`},
  journal = {Journal of Open Source Software}
}

```

When describing or discussing the theory underlying the `cvCovEst` method, or
simply using the method, please cite the pre-print below.

```
@misc{boileau2021,
      title={Cross-Validated Loss-Based Covariance Matrix Estimator Selection in High Dimensions}, 
      author={Philippe Boileau and Nima S. Hejazi and Mark J. van der Laan and Sandrine Dudoit},
      year={2021},
      eprint={2102.09715},
      archivePrefix={arXiv},
      primaryClass={stat.ME}
}
```

---

## License

&copy; 2020-2022 [Philippe Boileau](https://pboileau.ca)

The contents of this repository are distributed under the MIT license. See file
[`LICENSE.md`](https://github.com/PhilBoileau/cvCovEst/blob/master/LICENSE.md)
for details.

---

## References
