---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R/`cvCovEst`

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/PhilBoileau/cvCovEst.svg?token=YL3L6rYQtszHibWHgFxU&branch=master)](https://travis-ci.com/PhilBoileau/cvCovEst)
[![Codecov test coverage](https://codecov.io/gh/PhilBoileau/cvCovEst/branch/master/graph/badge.svg)](https://codecov.io/gh/PhilBoileau/cvCovEst?branch=master)
<!-- badges: end -->

> Cross-Validated Covariance Matrix Estimation

__Authors:__ [Philippe Boileau](https://pboileau.ca) and
  [Nima Hejazi](https://nimahejazi.org)

---

## What's `cvCovEst`?

`cvCovEst` implements an efficient cross-validated procedure for covariance
matrix estimation, particularly useful in high-dimensional settings. The general
methodology allows for cross-validation to be used to data adaptively identify
the optimal estimator of the covariance matrix from a pre-specified set of
candidate estimators. Also included are diagnostic tools for assessing the
quality of covariance matrix estimators.

---

## Installation

Install the _development version_ of the `cvCovEst` package from GitHub via
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval=FALSE}
remotes::install_github("PhilBoileau/cvCovEst")
```

---

## Example

To illustrate how `cvCovEst` may be used to select an optimal covariance matrix
estimator via cross-validation, consider the following toy example:

```{r example, message=FALSE, warning=FALSE}
library(MASS)
library(cvCovEst)
set.seed(1584)

# generate a 50x50 covariance matrix with unit variances and off-diagonal
# elements equal to 0.5
Sigma <- matrix(0.5, nrow = 50, ncol = 50) + diag(0.5, nrow = 50)

# sample 200 observations from multivariate normal with mean = 0, var = Sigma
dat <- mvrnorm(n = 200, mu = rep(0, 50), Sigma = Sigma)

# run CV-selector
cv_cov_est_out <- cvCovEst(
    dat = dat,
    estimators = c("linearShrinkEst", "thresholdingEst"),
    estimator_params = list("linearShrinkEst" = list("alpha" = c(0.1, 0.9)),
                            "thresholdingEst" = list("gamma" = c(0.2, 2))),
    cv_scheme = "v_fold", mc_split = 0.5,
    v_folds = 5, boot_iter = 20,
    center = TRUE, scale = FALSE, parallel = FALSE
  )

# print the table of risk estimates
# NOTE: the estimated covariance matrix is accessible via the `$estimate` slot
cv_cov_est_out$risk_df
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/PhilBoileau/cvCovEst/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/PhilBoileau/cvCovEst/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## License

&copy; 2020 [Philippe Boileau](https://pboileau.ca)

The contents of this repository are distributed under the MIT license. See file
[`LICENSE.md`](https://github.com/PhilBoileau/cvCovEst/blob/master/LICENSE.md)
for details.
