---
title: "`cvCovEst`: Cross-Validated Covariance Matrix Estimation"
author: "[Philippe Boileau](https://pboileau.ca), [Brian Collica](https://github.com/bcollica), [Nima
  Hejazi](https://nimahejazi.org)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{cvCovEst: Cross-Validated Covariance Matrix Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Background and Motivation

When the number of observations in a dataset far exceeds the number of features,
the estimator of choice for the covariance matrix is the sample covariance
matrix. It is an efficient estimator under minimal regularity assumptions on the
data-generating distribution. In high-dimensional regimes, however, this
estimator leaves much to be desired: The sample covariance matrix is either
singular, numerically unstable, or both, thereby amplifying estimation error.

As high-dimensional data have become widespread, researchers have derived many
novel covariance matrix estimators to remediate the sample covariance matrix's
deficiencies. These estimators come in many flavours, though most are
constructed by regularizing the sample covariance matrix, or through the
estimation of latent factors. A comprehensive review is provided by @fan2016.

This variety brings with it many challenges. Identifying an "optimal" estimator
from among a collection of candidates can prove a daunting task, one whose
objectivity is often compromised by the analyst's decisions.  Though data-driven
approaches for selecting an optimal estimator from among estimators belonging to
certain (limited) classes have been derived, the question of selecting an
estimator from among a diverse collection of candidates remains unaddressed.

We therefore offer a general, cross-validation-based framework for
covariance matrix estimator selection to tackle just that. The
high-dimensional asymptotic optimality of selections are guaranteed based upon
extensions of the seminal work of @laan_dudoit:2003, @dudoit2005, and @vaart2006
on data-adaptive estimator selection to high-dimensional covariance matrix
estimation [@boileau2021]. The interested reader is invited to review
theoretical underpinnings of the methodology as described in @boileau2021.

## Cross-validated Covariance Matrix Estimation

Let there be a high-dimensional dataset comprising $n$ realizations of $i.i.d.$
$p$-length random vectors with a possibly nonparametric data-generating
distribution. Our goal is to estimate these random vectors' covariance matrix,
which may be accomplished using our general cross-validated estimator selection
framework.

Given a library of candidate estimators, a loss function, and a choice of
cross-validation scheme, `cvCovEst` will identify the asymptotically optimal
estimator of the covariance matrix from among all candidates. It subsequently
estimates this parameter using the selected candidate. An example is provided
below. Lists and brief descriptions of implemented candidate estimators, loss
functions, and cross-validation schemes are provided in the sequel.

```{r brief-example}
library(MASS)
library(cvCovEst)
set.seed(1584)

# generate a 50x50 covariance matrix with unit variances and off-diagonal
# elements equal to 0.5
sigma <- matrix(0.5, nrow = 50, ncol = 50) + diag(0.5, nrow = 50)

# sample 50 observations from multivariate normal with mean = 0, var = Sigma
dat <- mvrnorm(n = 50, mu = rep(0, 50), Sigma = sigma)

# run CV-selector
cv_cov_est_out <- cvCovEst(
    dat = dat,
    estimators = c(
      linearShrinkLWEst, denseLinearShrinkEst,
      thresholdingEst, poetEst, sampleCovEst
    ),
    estimator_params = list(
      thresholdingEst = list(gamma = c(0.2, 0.4)),
      poetEst = list(lambda = c(0.1, 0.2), k = c(1L, 2L))
    ),
    cv_loss = cvMatrixFrobeniusLoss,
    cv_scheme = "v_fold",
    v_folds = 5,
  )

# print the table of risk estimates
cv_cov_est_out$risk_df

# print a subset of the selected estimator's estimate
cv_cov_est_out$estimate[1:5, 1:5]
```

### Candidate Estimators

Covariance matrix estimators implemented in the `cvCovEst` package are
catalogued in the following table. These estimators are fed to the `cvCovEst`
function through the `estimators` argument as a vector. If these estimators rely
on hyperparameters, then they must be passed to the `estimator_params` as a
list. Depending on one's assumptions --- or lack thereof --- about the true
covariance matrix, one may choose to use a subset of these estimators or all of
them. Of course, they may also be used as standalone functions.

|Estimator | Implementation | Description |
|----------|----------|-------------|
| Sample covariance matrix | `sampleCovEst` | The sample covariance matrix. |
| Hard thresholding [@Bickel2008_thresh] | `thresholdingEst` | Applies a hard thresholding operator to the entries of the sample covariance matrix |
| SCAD thresholding [@rothman2009;@fan2001] | `scadEst` | Applies the SCAD thresholding operator to the entries of the sample covariance matrix.|
| Adaptive LASSO [@rothman2009] | `adaptiveLassoEst` | Applies the adaptive LASSO thresholding operator to the entries of the sample covariance matrix |
| Banding [@bickel2008_banding] | `bandingEst` | Replaces the sample covariance matrix's off-diagonal bands by zeros. |
| Tapering [@cai2010] | `taperingEst` | Tapers the sample covariance matrix's off-diagonal bands, eventually replacing them by zeros. |
| Optimal Linear Shrinkage [@Ledoit2004] | `linearShrinkLWEst` | Asymptotically optimal shrinkage of the sample covariance matrix towards the identity. |
| Linear Shrinkage [@Ledoit2004] | `linearShrinkEst` | Shrinkage of the sample covariance matrix towards the identity, but the shrinkage is controlled by a hyperparameter. |
| Dense Linear Shrinkage [@shafer2005] | `denseLinearShrinkEst` | Asymptotically optimal shrinkage of the sample covariance matrix towards a dense matrix whose diagonal elements are the mean of the sample covariance matrix's diagonal, and whose off-diagonal elements are the mean of the sample covariance matrix's off-diagonal elements. |
| Nonlinear Shrinkage [@Ledoit2020] | `nlShrinkLWEst` | Analytical estimator for the nonlinear shrinkage of the sample covariance matrix. |
| POET [@fan2013] | `poetEst` | An estimator based on latent variable estimation and thresholding. |
| Robust POET [@fan2018] | `robustPoetEst` | A robust (and more computationally taxing) take on the POET estimator. |

Note that `cvCovEst` only functions with estimators native to this package. If
you'd like to request a new estimator implementation, please submit an
[issue](https://github.com/PhilBoileau/cvCovEst/issues) to the queue.

### Loss Functions

Given a collection of candidate estimators, `cvCovEst` compares their
conditional cross-validated risks to identify the optimal selection. The loss
function used to compute these risks should reflect both aspects of the
data-generating distribution and the goal of the estimation procedure. This
package currently implements three loss functions:

| Loss | Implementation | Description |
|------|----------------|-------------|
| Matrix-based Frobenius | `cvMatrixFrobeniusLoss` | The default, based on the Frobenius norm. Appropriate when the dataset's features are of similar magnitudes. |
| Variance-scaled matrix-based Frobenius | `cvScaledMatrixFrobeniusLoss` | A scaled version of the matrix-based Frobenius loss, where weights are the inverse of products from the sample covariance matrix's diagonal. Appropriate when the features of the dataset are of different magnitudes. |
| Observation-based Frobenius | `cvFrobeniusLoss` | Based on the Frobenius norm and the rank-1, observation-level estimates of the sample covariance matrix. Its selections are equivalent to that of the matrix-based Frobenius, though less computationally efficient. However, the optimality results of @boileau2021 rely on it. |

The choice of loss function is set trough the `cv_loss` argument. Like the
candidate estimators, `cvCovEst` only supports loss functions implemented in
this package. Please submit suggestions to the
[issue](https://github.com/PhilBoileau/cvCovEst/issues) queue.

### Supported Cross-Validation Schemes

Two cross-validation schemes are currently supported by `cvCovEst`. Please
consider filing an [issue](https://github.com/PhilBoileau/cvCovEst/issues) in
the queue to request the implementation of another cross-validation scheme.

| Scheme | Details |
|--------|---------|
| V-fold | To use V-fold cross-validation, set the `cv_scheme` argument in `cvCovEst` to `"v_fold"`, and set the number of folds through the `v_folds` argument. `cvCovEst` defaults to 5-fold cross-validation. |
| Monte-Carlo | To perform Monte-Carlo cross-validation, set the `cv_scheme` argument to `"mc"`. Set the proportion of data to be used in each validation set using the `mc_split` argument, and the number of iterations to perform with the `v_folds` argument. |


## Diagnostic Tools

In addition to selecting an optimal estimator, the `cvCovEst` package contains
summary and plotting methods which highlight the statistical properties of the
candidate estimators, and inform the performance of the selection framework.
These tools help build intuition about these estimators' behavior and
allows for the evaluation of their performance over varying inputs.

### Summary Method 

The `summary` method for `cvCovEst` accepts an `object` argument, a named
`list` of class `cvCovEst`, and a `summ_fun` argument, a character vector
specifying the type of summary function to use.

These summary functions allow the user to quickly compare the performance of
several classes of estimators and compute other metrics of interest.  The
choices of `summ_fun` and their outputs are described below:

| Summary | Implementation | Description |
|---------|----------------|-------------|
| Empirical Risk by Estimator Class | `empRiskByClass` | Returns the minimum, 1^st^ quartile, median, 3^rd^ quartile, and maximum of the empirical risk associated with each class of estimator passed to `cvCovEst`. |
| Best Performing Estimator by Class | `bestInClass` | Returns the specific hyperparameters, if applicable, of the best performing estimator within each class. |
| Worst Performing Estimator by Class | `worstInClass` | Returns the specific hyperparameters, if applicable, of the worst performing estimator within each class. | 
| Empirical Risk by Hyperparameter | `hyperRisk` | For estimators that take hyperparameters as arguments, this returns the hyperparameters associated with the minimum, 1^st^ quartile, median, 3^rd^ quartile, and maximum of the empirical risk within each class of estimator.  Each class has its own `tibble` which are returned as a `list`. |

### Plot Method

The `plot` method for `cvCovEst` allows users to visualize three main
`plot_type`s of the candidate estimators: covariance heat maps (`heatmap`),
eigenvalue plots (`eigen`), and the empirical risk (`risk`) as a function of
the hyperparameters (for applicable estimator classes).  If users do not
specify a plot type, then all three plots are combined into one figure for the
optimal estimator selected by `cvCovEst`.  Users can also achieve this by
setting `plot_type = "summary"`.

The heat maps and eigenvalue plots facilitate comparisons both within and
between estimator classes by allowing multiple values to be passed as
`estimator` and `stat` arguments.  

Additional arguments specific to each `plot_type` are outlined below:

#### Heat Map

| Argument | Description |
|----------|-------------|
| abs_v | If `TRUE`, then the absolute value of the covariance is mapped.  Otherwise, the signed value is used. |

#### Eigenvalue Plot

| Argument | Description |
|----------|-------------|
| leading | If `TRUE`, then the k leading eigenvalues are displayed.  Otherwise, the k trailing eigenvalues are displayed. | 
| k | The number of leading or trailing eigenvalues to plot. |

#### Empirical Risk

These two additional arguments only apply to estimators with multiple hyperparameters:

| Argument | Description |
|----------|-------------|
| switch_vars | If `TRUE`, the hyperparameters used for the x-axis and factor variables are switched. |
| min_max | If `TRUE`, only the minimum and maximum values of the factor hyperparameter will be used. |


# Simulated Data Analysis Example

To show how the plot and summary methods can be used, data is simulated from a
predetermined covariance matrix following a Toeplitz structure.  The data is
then passed to `cvCovEst` along with a handful of estimators. 

```{r toep-sim}
set.seed(1584)

toep_sim <- function(p, rho, alpha) {
    times <- seq_len(p)
    H <- abs(outer(times, times, "-")) + diag(p)
    H <- H^-(1 + alpha) * rho
    covmat <- H + diag(p) * (1 - rho)

    sign_mat <- sapply(
      times,
      function(i) {
        sapply(
          times,
          function(j) {
            (-1)^(abs(i - j))
          }
        )
      }
    )
    return(covmat * sign_mat)
}

# simulate a 100 x 100 covariance matrix
sim_covmat <- toep_sim(p = 100, rho = 0.6, alpha = 0.3)

# sample 75 observations from multivariate normal mean = 0, var = sim_covmat
sim_dat <-  MASS::mvrnorm(n = 75, mu = rep(0, 100), Sigma = sim_covmat)

# run CV-selector
cv_cov_est_sim <- cvCovEst(
  dat = sim_dat,
  estimators = c(
    linearShrinkEst, thresholdingEst, bandingEst, adaptiveLassoEst,
    sampleCovEst, taperingEst
  ),
  estimator_params = list(
    linearShrinkEst = list(alpha = seq(0.25, 0.75, 0.05)),
    thresholdingEst = list(gamma = seq(0.25, 0.75, 0.05)),
    bandingEst = list(k = seq(2L, 10L, 2L)),
    adaptiveLassoEst = list(lambda = c(0.1, 0.25, 0.5, 0.75, 1), n = seq(1, 5)),
    taperingEst = list(k = seq(2L, 10L, 2L))
  ),
  cv_scheme = "v_fold",
  v_folds = 5
)
```


### Summary Method 

The `summary` method is then used to compare the best performing estimators in
each class:

```{r summary-sim}
cv_sum <- summary(cv_cov_est_sim, dat_orig = sim_dat)
cv_sum$bestInClass
```

In this case, the tapering estimator with `k = 4` achieves the lowest empirical
risk.  It is also one of the only positive-definite estimators along with the linear shrinkage estimator, although linear shrinkage yields a better conditioned estimator in this case.  We can take a closer look at the estimator's performance based on other
possible hyperparameter values by hashing `taperingEst` from the `hyperRisk`
list.

```{r hyperRisk-sim}
cv_sum$hyperRisk$bandingEst
```


### Plot Method

By specifying `plot_type = "risk"`, we can see the change in empirical risk as
the value of `k` changes.  A summary of the cross-validation scheme and loss
function is displayed at the bottom of all `cvCovEst` plot outputs.   


```{r risk-sim, fig.height = 3.5, fig.width=4}
plot(cv_cov_est_sim, dat_orig = sim_dat, plot_type = "risk")
```

We can also examine the matrix structure as the value of `k` changes.
Examining the overall sparsity of the resulting estimator can be useful since,
in some cases, the assumption of sparsity is not warranted. Note that the
absolute values of the estimate's entries are displayed to emphasize the
structural differences between choices of hyperparameters.


```{r multi-heat-sim, fig.height=3.5, fig.width=7}
plot(cv_cov_est_sim, dat_orig = sim_dat, plot_type = "heatmap",
     stat = c("min", "median", "max"))
```

If the signs of the covariances are of interest, they can be displayed by
setting `abs_v = FALSE`:

```{r sign-multi-heat-sim, fig.height=3.5, fig.width=7}
plot(cv_cov_est_sim, dat_orig = sim_dat, plot_type = "heatmap",
     stat = c("min", "median", "max"), abs_v = FALSE)
```

The difference between the optimal estimator selected by `cvCovEst` and the
sample covariance matrix is clear when displaying their respective heat maps
side by side.


```{r samp-heat-sim, fig.height=3.5, fig.width=5}
plot(cv_cov_est_sim, dat_orig = sim_dat, plot_type = "heatmap",
     estimator = c("bandingEst", "sampleCovEst"),
     stat = c("min"), abs_v = FALSE)
```

We may also be interested in the eigenvalues of the various banding estimators.
The distribution of eigenvalues relays information such as the condition number
or the positive-definiteness of the resulting estimator.  As with the other
plot types, if the `estimator` argument is not specified, the default is to
display the optimal estimator selected by `cvCovEst`.


```{r eigen-sim, fig.height = 3.5, fig.width=5}
plot(cv_cov_est_sim, dat_orig = sim_dat, plot_type = "eigen",
     stat = c("min", "median", "max"))
```

Specifying multiple values in the `estimator` argument allows us compare the
eigenvalues of the other estimator classes as well.

```{r multi-eigen-sim, fig.height=4.5, fig.width=6}
plot(cv_cov_est_sim, dat_orig = sim_dat, plot_type = "eigen",
     stat = c("min", "median", "max"),
     estimator = c("bandingEst", "thresholdingEst", "linearShrinkEst",
                    "adaptiveLassoEst"))
```


As previously mentioned, simply calling the `plot` method on the output of
`cvCovEst` and providing the original data will result in a visual summary of
the selected estimator.  

```{r plot-summary, fig.height=7, fig.width=7.5}
plot(cv_cov_est_sim, dat_orig = sim_dat)
```

## References
